Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-21 RevisedSubmission/hines_hetland_2017_gji.tex
justified text file

\begin{summary}
Transient strain rates derived from GNSS data can be used to detect and understand geophysical phenomena such as slow slip events and postseismic deformation. Here we propose using Gaussian process regression (GPR) as a tool for estimating transient strain rates from GNSS data. GPR is a non-parametric, Bayesian method for interpolating scattered data. Transient strain rates estimated with GPR have meaningful uncertainties, allowing geophysical signal to be easily discerned from noise. In our approach, we assume a stochastic prior model for transient displacements. The prior describes how much one expects transient displacements to covary spatially and temporally. A posterior estimate of transient strain rates is obtained by differentiating the posterior displacements. One limitation with GPR is that it is not robust against outliers, so we introduce a pre-processing method for detecting and removing outliers in GNSS data. As a demonstration, we use GPR to detect transient strain resulting from slow slip events in the Pacific Northwest. Maximum likelihood methods are used to constrain a prior model for transient displacements in this region. The temporal covariance of our prior model is described by a compact Wendland covariance function, which significantly reduces the computational burden that can be associated with GPR. Our results reveal the spatial and temporal evolution of strain from slow slip events. We verify that the transient strain estimated with GPR is in fact geophysical signal by comparing it to the seismic tremor that is associated with Pacific Northwest slow slip events.
\end{summary}

@@

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates estimated from GNSS data have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Transient crustal strain, which is caused by geophysical phenomena such as slow slip events (SSEs) or postseismic deformation, is also relevant for assessing seismic hazard. While transient strain itself is not damaging, there is a risk that it can trigger major earthquakes \citep{Roeloffs2006,Freed2001}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it possible to identify transient strain with high fidelity. Developing and improving upon methods for deriving secular and transient strain from GNSS data is an active area of research.

@@

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a continuous stochastic process whose value at any collection of points can be described as a Gaussian random vector. Just as a Gaussian random vector is fully determined by a mean vector and a covariance matrix, a Gaussian process is fully determined by a mean function and a covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has the mean function $\mathrm{E}[B(t)] = 0$ and the covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If there is no information available to help choose an appropriate prior Gaussian process, then maximum likelihood methods can be used to objectively choose one that is most consistent with the data. Once a prior is selected, we incorporate GNSS observations with the prior to form a posterior estimate of transient strain. The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, which are constrained by GNSS data.

We let $\vec{u}(p) = (u_\mathrm{e}(p),u_\mathrm{n}(p))$ be our prior understanding of transient displacements. Since transient displacements are not precisely known, we do not consider $\vec{u}$ to be a deterministic function. Instead, $\vec{u}$ is a stochastic process containing a distribution of functions that could potentially describe transient displacements. We let each component of $\vec{u}$ be a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we write our prior on each component of $\vec{u}$ as $u_i \sim \mathcal{GP}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite in order to be a valid covariance function. By definition, $C_{u_i}$ is a positive definite function if the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures that $\vec{u}$ is spatially and temporally differentiable, allowing us to compute $\dot\varepsilon$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec. 10A]{Papoulis1991} for a definition of stochastic differentiation and the conditions for differentiability).

To give an example of $u_i$, we can use a squared exponential function for $X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\vec{x},\vec{x}') = \exp\left(\frac{-||\vec{x} - \vec{x}'||_2^2}{2\ell^2}\right), \ \ \ \ T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and differentiability. The parameters $\ell$ and $\tau$ control the length-scale and time-scale, respectively, of realizations (i.e., sample functions) of $u_i$. The parameter $\phi$, which we have arbitrarily chosen to incorporate into $T_i$ rather than $X_i$, controls the amplitude of realizations of $u_i$. Ideally, we want realizations of $u_i$ to have a length-scale, time-scale, and amplitude that resembles what we expect from true signal. A few realizations for $u_i$ are shown in Figure XXX. While the squared exponential function is a commonly used covariance function for GPR, it is not appropriate for every application. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe. To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t^{(k)}$. We describe this observation, $d_i^{*(jk)}$, as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_i^{(jk)} = &u_i\left(\vec{x}^{(j)},t^{(k)}\right) + 
              \eta_i^{(jk)} + 
              a_i^{(j)} + b_i^{(j)}t^{(k)} + \\
             &c_i^{(j)}\sin\left(2 \pi t^{(k)}\right) +  
              e_i^{(j)}\cos\left(2 \pi t^{(k)}\right) + \\ 
             &f_i^{(j)}\sin\left(4 \pi t^{(k)}\right)  + 
              g_i^{(j)}\cos\left(4 \pi t^{(k)}\right), 
\end{split}
\end{align}
where $\eta_i^{(jk)}$ describes noise, $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t^{(k)}$). We then consider the column vector of $n$ GNSS displacement observations, $\mitbf{d}^*_i$, where the subscript indicates that we are still only considering component $i$ of displacements. The observations are made at $m$ stations, and the times and positions for each observation are described by the set $\mitbf{P}$. $\mitbf{d}^*_i$ can be considered a realization of the random vector $\mitbf{d}_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each point in $\mitbf{P}$. To write out $\mitbf{d}_i$ explicitly, we let $\mitbf{G}$ be an $n \times 6m$ matrix consisting of the basis functions from eq. (\ref{eq:Data}) (i.e., the linear trends and sinusoids for each station) evaluated at each point in $\mitbf{P}$. The coefficients corresponding to each basis function are collected into the column vector $\mitbf{m}_i$. The noise for all the observations are described by the column vector $\mitbf{\eta}_i$. We can then write $\mitbf{d}_i$ as 
\begin{equation}\label{eq:DataVec}
\mitbf{d}_i = u_i(\mitbf{P}) + \mitbf{\eta}_i + \mitbf{G}\mitbf{m}_i,
\end{equation}
where the notation $u_i(\mitbf{P})$ represents the column vector $[u_i(p)]_{p \in \mitbf{P}}$.

We assume a diffuse prior for the components of $\mitbf{m}_i$, that is to say $\mitbf{m}_i \sim \mathcal{N}(\mitbf{0},\kappa^2\mitbf{I})$ in the limit as $\kappa \to \infty$. Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study. We assume that $\mitbf{\eta}_i$ is a spatially and/or temporally correlated random vector distributed as $\mathcal{N}(\mitbf{0},\mitbf{C}_{\eta_i})$. For example, $\mitbf{\eta}_i$ can be uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. The appropriate noise model may vary depending on the application, and we hold off on specifying the covariance matrix, $\mitbf{C}_{\eta_i}$, until Section \ref{sec:NoiseModel}. We are now able to write the distribution of $\mitbf{d}_i$ as 
\begin{equation}\label{eq:DataDist}
\mitbf{d}_i \sim \mathcal{N}\left(\mitbf{0}, C_{u_i}(\mitbf{P},\mitbf{P}) + \mitbf{C}_{\eta_i}  + \kappa^2\mitbf{G}\mitbf{G}^T \right),
\end{equation}
where $C_{u_i}(\mitbf{P},\mitbf{P})$ represents the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$.

We form a posterior estimate of transient displacements, denoted as $\hat{u}_i$, by updating $u_i$ with the fact that $\mitbf{d}_i^*$ was realized from the random vector $\mitbf{d}_i$, that is to say $\hat{u}_i = u_i |(\mitbf{d}_i = \mitbf{d}^*_i)$. A general solution for $\hat{u}_i$ is derived in \citet[sec. 8.9]{VonMises1964}, where we find that $\hat{u}_i$ is distributed as $\mathcal{GP}(\mu_{\hat{u}_i},C_{\hat{u}_i})$ with mean function

@@

\mu_{\hat{u}_i}(p) &= \mathrm{E}\left[u_i(p)\right] + 
                      \mathrm{Cov}\left[u_i(p),\mitbf{d}_i\right] 
                      \mathrm{Cov}\left[\mitbf{d}_i\right]^{-1}
                      \left(\mitbf{d}^*_i - \mathrm{E}\left[\mitbf{d}_i\right]\right)\nonumber \\
                   &= C_{u_i}(p,\mitbf{P})\left( C_{u_i}(\mitbf{P},\mitbf{P}) + \mitbf{C}_{\eta_i} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}\mitbf{d}^*_i

@@

C_{\hat{u}_i}(p,p') &= \mathrm{Cov}\left[u_i(p),u_i(p')\right] - 
                       \mathrm{Cov}\left[u_i(p),\mitbf{d}_i\right] 
                       \mathrm{Cov}\left[\mitbf{d}_i\right]^{-1}
                       \mathrm{Cov}\left[\mitbf{d}_i,u_i(p')\right]\nonumber \\
                    &= C_{u_i}(p,p') - C_{u_i}(p,\mitbf{P})\left( C_{u_i}(\mitbf{P},\mitbf{P}) + \mitbf{C}_{\eta_i} + \kappa^2\mitbf{G}\mitbf{G}^T \right)^{-1}C_{u_i}(\mitbf{P},p').

@@

However, we are interested in the limit as $\kappa \to \infty$, and the form for eqs. (\ref{eq:PosteriorMean}) and (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We use a partitioned matrix inversion identity \citep[sec. 2.7.4]{Press2007} to rewrite eqs. (\ref{eq:PosteriorMean}) and (\ref{eq:PosteriorCov}) as

@@

\mu_{\hat{u}_i}(p) = \left[\begin{array}{cc}
                         C_{u_i}(p,\mitbf{P}) & \mitbf{0} \\

@@

                         C_{u_i}(\mitbf{P},\mitbf{P}) + \mitbf{C}_{\eta_i} & \mitbf{G} \\
                         \mitbf{G}^T  & -\kappa^{-2} \mitbf{I} \\

@@

                         \mitbf{d}^*_i \\
                         \mitbf{0} \\

@@

C_{\hat{u}_i}(p,p') = C_{u_i}(p,p') - \\ 

@@

                          C_{u_i}(p,\mitbf{P}) & \mitbf{0} \\

@@

                          C_{u_i}(\mitbf{P},p') \\
                          \mitbf{0} \\

@@

Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}_i$,

@@

                         C_{u_i}(\mitbf{P},\mitbf{P}) + \mitbf{C}_{\eta_i} & \mitbf{G} \\
                         \mitbf{G}^T  & \mitbf{0} \\

@@

C_{\hat{u}_i}(p,p') = C_{u_i}(p,p') - 

@@

To ensure that the inverse matrices in eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) exist, each column in $\mitbf{G}$ must be linearly independent. This condition tends to be violated when there are too few observations at a station. In that case, a singular value decomposition can be used to remove linearly dependent components from $\mitbf{G}$.

It should be noted that we have ignored any covariances between the easting and northing components of $\vec{u}$ and $\vec{\mitbf{d}}$. This simplification reduces the computational complexity of evaluating the posterior transient displacements because each component can be evaluated independently. However, we are inherently assuming that the principle axes describing the distribution of $\vec{u}(p)$ and $\vec{d}^{(jk)}$ are aligned with the cardinal directions, which is admittedly an arbitrary assumption.

The posterior transient displacements are spatially and temporally continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, $\hat{u}_i$ is spatially and temporally differentiable, allowing us to formulate $\dot\varepsilon$ at any $p$ that we may be interested in. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i} \right).
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $\hat{u}_i$ and $\hat{u}_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. From \citet[sec. 10]{Papoulis1991}, we find that the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are 
\begin{align}\label{eq:StrainMean}
\mu_{\dot\varepsilon_\mathrm{ee}}(p) &= \frac{\partial^2 \mu_{\hat{u}_\mathrm{e}}(p)}{\partial t \, \partial x_\mathrm{e}} \\
\mu_{\dot\varepsilon_\mathrm{nn}}(p) &= \frac{\partial^2 \mu_{\hat{u}_\mathrm{n}}(p)}{\partial t \, \partial x_\mathrm{n}} \\

>>>>

%% DEFINE MATH MODE MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pos}{\vec{x}} % position vector
\newcommand{\data}{\mitbf{d}} % data vector
\newcommand{\post}{\hat{u}} % posterior
\newcommand{\points}{\mitbf{P}} % observation points
\newcommand{\strain}{\dot\varepsilon} % strain rate
\newcommand{\E}[1]{\e\left[ #1 \right]} % Expected value
\newcommand{\Cov}[1]{\mathrm{Cov}\left[ #1 \right]} % Expected value
\newcommand{\e}{\mathrm{e}} % east subscript
\newcommand{\n}{\mathrm{n}} % north subscript
\newcommand{\G}{\mitbf{G}} % basis functions
\newcommand{\zeros}{\mitbf{0}} % zeros matrix
\newcommand{\eye}{\mitbf{I}} % identity matrix

%% SUMMARY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{summary}

Transient strain rates derived from GNSS data can be used to detect
and understand geophysical phenomena such as slow slip events and
postseismic deformation. Here we propose using Gaussian process
regression (GPR) as a tool for estimating transient strain rates from
GNSS data. GPR is a non-parametric, Bayesian method for interpolating
scattered data. Transient strain rates estimated with GPR have
meaningful uncertainties, allowing geophysical signal to be easily
discerned from noise. In our approach, we assume a stochastic prior
model for transient displacements. The prior describes how much one
expects transient displacements to covary spatially and temporally. A
posterior estimate of transient strain rates is obtained by
differentiating the posterior displacements. One limitation with GPR
is that it is not robust against outliers, so we introduce a
pre-processing method for detecting and removing outliers in GNSS
data. As a demonstration, we use GPR to detect transient strain
resulting from slow slip events in the Pacific Northwest. Maximum
likelihood methods are used to constrain a prior model for transient
displacements in this region. The temporal covariance of our prior
model is described by a compact Wendland covariance function, which
significantly reduces the computational burden that can be associated
with GPR. Our results reveal the spatial and temporal evolution of
strain from slow slip events. We verify that the transient strain
estimated with GPR is in fact geophysical signal by comparing it to
the seismic tremor that is associated with Pacific Northwest slow slip
events.

\end{summary}

@@

%% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:Introduction}

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Consequently, secular crustal
strain rates estimated from GNSS data have been used to constrain
seismic hazard models such as UCERF3 \citep{Field2014}. Transient
crustal strain, which is caused by geophysical phenomena such as slow
slip events (SSEs) or postseismic deformation, is also relevant for
assessing seismic hazard. While transient strain itself is not
damaging, there is a risk that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}. Dense networks of continuous GNSS
stations, such as the Plate Boundary Observatory (PBO), make it
possible to identify transient strain with high fidelity. Developing
and improving upon methods for deriving secular and transient strain
from GNSS data is an active area of research.

@@

Here we propose using Gaussian process regression (GPR)
\citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR
is closely related to kriging \citep{Cressie1992} and least squares
collocation \citep{Moritz1978}. The latter has been used by
\citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain
rates from GNSS data. GPR is a Bayesian, non-parametric method for
inferring a continuous signal from scattered data. Since GNSS stations
are irregularly spaced and observation times may differ between
stations, GPR is an ideal tool for synthesizing discrete GNSS data
into a spatially and temporally continuous representation of surface
deformation. GPR is Bayesian in that we describe our prior
understanding of the geophysical signal with a Gaussian process. A
Gaussian process is a continuous stochastic process whose value at any
collection of points can be described as a Gaussian random vector.
Just as a Gaussian random vector is fully determined by a mean vector
and a covariance matrix, a Gaussian process is fully determined by a
mean function and a covariance function. For example, Brownian motion,
$B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has
the mean function $\E{B(t)} = 0$ and the covariance function
$\Cov{B(t),B(t')} = \min(t,t')$, where $t,t' \ge 0$. If there is no
information available to help choose an appropriate prior Gaussian
process, then maximum likelihood methods can be used to objectively
choose one that is most consistent with the data. Once a prior is
selected, we incorporate GNSS observations with the prior to form a
posterior estimate of transient strain. The posterior transient strain
is also a Gaussian process, and we can use its distribution to
confidently discern geophysical signal from noise. We use GPR to infer
strain resulting from SSEs in the Pacific Northwest, demonstrating
that GPR is an effective tool for detecting transient geophysical
processes.

%% ESTIMATING TRANSIENT STRAIN RATES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Estimating Transient Strain Rates}\label{sec:Method}

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D map projection that is
aligned with the cardinal directions. We determine $\strain$ by
spatially and temporally differentiating estimates of transient
displacements, which are constrained by GNSS data.

We let $\vec{u}(p) = (u_\e(p),u_\n(p))$ be our prior understanding of
transient displacements. Since transient displacements are not
precisely known, we do not consider $\vec{u}$ to be a deterministic
function. Instead, $\vec{u}$ is a stochastic process containing a
distribution of functions that could potentially describe transient
displacements. We let each component of $\vec{u}$ be a Gaussian
process with zero mean and a generic covariance function
$\Cov{u_i(p),u_i(p')} = C_{u_i}(p,p')$. Using a more concise notation,
we write our prior on each component of $\vec{u}$ as $u_i \sim
\mathcal{GP}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be
positive definite in order to be a valid covariance function. By
definition, $C_{u_i}$ is a positive definite function if the matrix
$[C_{u_i}(p,p')]_{(p,p') \in \points \times \points}$ is positive
definite for any set of points $\points$ \citep[sec.
2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into
spatial and temporal functions as
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\pos,t),(\pos',t')\right) 
              = X_i(\pos,\pos')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite,
$C_{u_i}$ is guaranteed to also be positive definite \citep[sec.
4.2.4]{Rasmussen2006}. We also require that the derivatives
$\partial^2 X_i(\pos,\pos')/ \partial x_j \partial x_j'$ and
$\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures
that $\vec{u}$ is spatially and temporally differentiable, allowing us
to compute $\strain$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec.
10A]{Papoulis1991} for a definition of stochastic differentiation and
the conditions for differentiability).

To give an example of $u_i$, we can use a squared exponential function
for $X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations (i.e.,
sample functions) of $u_i$. The parameter $\phi$, which we have
arbitrarily chosen to incorporate into $T_i$ rather than $X_i$,
controls the amplitude of realizations of $u_i$. Ideally, we want
realizations of $u_i$ to have a length-scale, time-scale, and
amplitude that resembles what we expect from true signal. A few
realizations for $u_i$ are shown in Figure XXX. While the squared
exponential function is a commonly used covariance function for GPR,
it is not appropriate for every application. The appropriate choice
for $X_i$ and $T_i$ may vary depending on the geophysical signal we
are trying to describe. To keep this section sufficiently general, we
hold off on specifying $X_i$ and $T_i$ until Section
\ref{sec:SignalModel}, where we estimate transient strain from slow
slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical
and non-physical processes that we are not interested in. We first
consider component $i$ of a single GNSS displacement observation made
at station $j$, which is located at $\pos^{(j)}$, and time $t^{(k)}$.
We describe this observation, $d_i^{*(jk)}$, as a realization of the
random variable
\begin{align}\label{eq:Data}
\begin{split}
d_i^{(jk)} = &u_i\left(\pos^{(j)},t^{(k)}\right) + 
              \eta_i^{(jk)} + 
              a_i^{(j)} + b_i^{(j)}t^{(k)} + \\
             &c_i^{(j)}\sin\left(2 \pi t^{(k)}\right) +  
              e_i^{(j)}\cos\left(2 \pi t^{(k)}\right) + \\ 
             &f_i^{(j)}\sin\left(4 \pi t^{(k)}\right)  + 
              g_i^{(j)}\cos\left(4 \pi t^{(k)}\right), 
\end{split}
\end{align}
where $\eta_i^{(jk)}$ describes noise, $a_i^{(j)}$ is an offset that
is unique for each station, $b_i^{(j)}$ is the secular velocity at
$\pos^{(j)}$, and the sinusoids describe seasonal deformation (using
units of years for $t^{(k)}$). We then consider the column vector of
$n$ GNSS displacement observations, $\data^*_i$, where the subscript
indicates that we are still only considering component $i$ of
displacements. The observations are made at $m$ stations, and the
times and positions for each observation are described by the set
$\points$. $\data^*_i$ can be considered a realization of the random
vector $\data_i$, which is formed by evaluating eq. (\ref{eq:Data}) at
each point in $\points$. To write out $\data_i$ explicitly, we let
$\G$ be an $n \times 6m$ matrix consisting of the basis functions from
eq. (\ref{eq:Data}) (i.e., the linear trends and sinusoids for each
station) evaluated at each point in $\points$. The coefficients
corresponding to each basis function are collected into the column
vector $\mitbf{m}_i$. The noise for all the observations are described
by the column vector $\mitbf{\eta}_i$. We can then write $\data_i$ as
\begin{equation}\label{eq:DataVec}
\data_i = u_i(\points) + \mitbf{\eta}_i + \G\mitbf{m}_i,
\end{equation}
where the notation $u_i(\points)$ represents the column vector
$[u_i(p)]_{p \in \points}$.

We assume a diffuse prior for the components of $\mitbf{m}_i$, that is
to say $\mitbf{m}_i \sim \mathcal{N}(\zeros,\kappa^2\eye)$ in the
limit as $\kappa \to \infty$. Of course, the secular velocities,
$b_i^{(j)}$, are spatially correlated and we could invoke a tectonic
model to form a prior on $b_i^{(j)}$. However, in our application to
the Pacific Northwest, we will be using displacement time series which
are long enough to sufficiently constrain $b_i^{(j)}$ for each
station, avoiding the need to incorporate a prior. Likewise, seasonal
deformation is spatially correlated \citep{Dong2002,Langbein2008}, and
it may be worth exploring and exploiting such a correlation in a
future study. We assume that $\mitbf{\eta}_i$ is a spatially and/or
temporally correlated random vector distributed as
$\mathcal{N}(\zeros,\mitbf{C}_{\eta_i})$. For example,
$\mitbf{\eta}_i$ can be uncorrelated white noise, temporally
correlated noise describing benchmark wobble
\citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated
noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. The
appropriate noise model may vary depending on the application, and we
hold off on specifying the covariance matrix, $\mitbf{C}_{\eta_i}$,
until Section \ref{sec:NoiseModel}. We are now able to write the
distribution of $\data_i$ as
\begin{equation}\label{eq:DataDist}
\data_i \sim \mathcal{N}\left(\zeros, C_{u_i}(\points,\points) + 
                                      \mitbf{C}_{\eta_i}  + 
                                      \kappa^2\G\G^T \right),
\end{equation}
where $C_{u_i}(\points,\points)$ represents the matrix
$[C_{u_i}(p,p')]_{(p,p') \in \points \times \points}$.

We form a posterior estimate of transient displacements, denoted as
$\post_i$, by updating $u_i$ with the fact that $\data_i^*$ was
realized from the random vector $\data_i$, that is to say $\post_i =
u_i |(\data_i = \data^*_i)$. A general solution for $\post_i$ is
derived in \citet[sec. 8.9]{VonMises1964}, where we find that
$\post_i$ is distributed as $\mathcal{GP}(\mu_{\post_i},C_{\post_i})$
with mean function

@@

\mu_{\post_i}(p) &= \E{u_i(p)} + 
                    \Cov{u_i(p),\data_i} 
                    \Cov{\data_i}^{-1}
                    \left(\data^*_i - \E{\data_i}\right)\nonumber \\
                 &= C_{u_i}(p,\points)\left( C_{u_i}(\points,\points) + 
                                             \mitbf{C}_{\eta_i} + 
                                             \kappa^2\G\G^T\right)^{-1}
                    \data^*_i

@@

C_{\post_i}(p,p') &= \Cov{u_i(p),u_i(p')} - 
                     \Cov{u_i(p),\data_i} 
                     \Cov{\data_i}^{-1}
                     \Cov{\data_i,u_i(p')} \nonumber \\
                  &= C_{u_i}(p,p') - 
                     C_{u_i}(p,\points)\left( C_{u_i}(\points,\points) + 
                                              \mitbf{C}_{\eta_i} + 
                                              \kappa^2\G\G^T \right)^{-1}
                     C_{u_i}(\points,p').

@@

However, we are interested in the limit as $\kappa \to \infty$, and
the form for eqs. (\ref{eq:PosteriorMean}) and (\ref{eq:PosteriorCov})
is not suitable for evaluating this limit. We use a partitioned matrix
inversion identity \citep[sec. 2.7.4]{Press2007} to rewrite eqs.
(\ref{eq:PosteriorMean}) and (\ref{eq:PosteriorCov}) as

@@

\mu_{\post_i}(p) = \left[\begin{array}{cc} 
                         C_{u_i}(p,\points) & \zeros \\

@@

                         C_{u_i}(\points,\points) + \mitbf{C}_{\eta_i} & \G \\
                         \G^T  & -\kappa^{-2} \eye \\

@@

                         \data^*_i \\
                         \zeros \\

@@

C_{\post_i}(p,p') = C_{u_i}(p,p') - \\ 

@@

                          C_{u_i}(p,\points) & \zeros \\

@@

                          C_{u_i}(\points,p') \\
                          \zeros \\

@@

Taking the limit as $\kappa \to \infty$, we get the solution for the
mean and covariance of $\post_i$,

@@

                         C_{u_i}(\points,\points) + \mitbf{C}_{\eta_i} & \G \\
                         \G^T  & \zeros \\

@@

C_{\post_i}(p,p') = C_{u_i}(p,p') - 

@@

To ensure that the inverse matrices in eqs. (\ref{eq:PosteriorMean3})
and (\ref{eq:PosteriorCov3}) exist, each column in $\G$ must be
linearly independent. This condition tends to be violated when there
are too few observations at a station. In that case, a singular value
decomposition can be used to remove linearly dependent components from
$\G$.

It should be noted that we have ignored any covariances between the
easting and northing components of $\vec{u}$ and $\vec{\data}$.
This simplification reduces the computational complexity of evaluating
the posterior transient displacements because each component can be
evaluated independently. However, we are inherently assuming that the
principle axes describing the distribution of $\vec{u}(p)$ and
$\vec{d}^{(jk)}$ are aligned with the cardinal directions, which is
admittedly an arbitrary assumption.
